---
title: "Processing of tuber bruising phenotype"
author: "Olivia Angelin-Bonnet"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
		      fig.align = 'center',
                      out.width = '\\linewidth',
                      fig.width = 12,
                      fig.height = 8)

here::i_am("genomics_gwas/5_gwas_phenotype_preprocessing.Rmd")

if (!require("readxl")) install.packages("readxl"); library(readxl)
if (!require("here")) install.packages("here"); library(here)
if(!require("bestNormalize")) install.packages("bestNormalize"); library(bestNormalize)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

```{r, cache = FALSE}
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5),
             legend.position = "bottom",
             text = element_text(size = 16))
```

## Data path

```{r}
samples_info_file = here("data/samples_info.csv")
bruising_file = here("data/raw_bruising_scores.csv")
```

```{r}
samples_info = read_csv(samples_info_file, col_types = cols())
```

## Reading the bruising phenotype

We'll read the bruising score dataset and get an average bruising score for each genotype (= average over 6 tubers, 3 from each biological replicate).

```{r}
bruising_df = read_csv(bruising_file, show_col_types = FALSE) %>% 
  select(Sample, Rep, starts_with("Bruise")) %>% 
  pivot_longer(cols = starts_with("Bruise"),
               names_to = "Bruise",
               values_to = "bruising_score") %>% 
  group_by(Sample) %>% 
  summarise(bruising_score_mean = mean(bruising_score, na.rm = TRUE))
```


We can assess whether each phenotype is normally distributed using the Shapiro-Wilk test.

```{r}
shapiro.test(bruising_df$bruising_score_mean)
```

The test shows evidence of non-normality.

## Normalisation of phenotypes using `bestNormalize`

The package `bestNormalize` tests different normalisation methods on a given dataset and automatically selects the best performing one. It can be used as follows:

```{r}
set.seed(1)
norm_vals <- bestNormalize(bruising_df$bruising_score_mean)
norm_vals
```


```{r}
tibble(Sample = bruising_df$Sample,
       bruising_score_mean = norm_vals$x.t) %>% 
  write_csv(here("genomics_gwas/processed_data/pheno_df_trans.csv"))
```


```{r}
shapiro.test(norm_vals$x.t)
```